---
title: "Markov Chain Monte Carlo"
---
## Why Markov Chain Monte Carlo

Markov Chain Monte Carlo allows us to sample from the probability distribution $\pi(x)$, which may not have been possible before.

::: {.callout-warning}
If you can sample exactly from the distribution, you do not need to use Markov Chain Monte Carlo.
For example, if you want to sample from a normal distribution, just sample from the normal distribution.
:::

## Markov Chains
::: {.callout title="Discrete Markov Chains"}
A discrete Markov Chain is a sequence of random variables that satisfy the Markov property.
That is, the next random variable in the sequence only depends upon the current state,
$$
X_{n+1} \sim P(\cdot | X_n)
$$
where $X_n$ is the $n$-th random variable in the sequence and $P(\cdot | X_n)$ is the proposal distribution.
:::

::: {.callout title="Detailed Balance"}
$$
\pi(x)p(x'|x) = \pi(x')p(x|x')
$$
where $\pi(x)$ is the stationary distribution of the Markov Chain and $p(x|x')$ is the transition probability from $x'$ to $x$.
The flow out of $x$ into $x'$ is balanced by the flow from $x'$ into $x$.
:::

It would be extremely helpful if we could easily construct the transition probabilities of a Markov Chain with a known stationary distribution.
If this was the case then we could generate a large number of samples from our Markov Chain and the stationary distribution of these samples will be samples from our distribution of interest.
This is the main premise behind Markov Chain Monte Carlo.

To construct the transition probabilities we decompose $p(x'|x)$ into an accept-reject step.
Namely,
$$
p(x'|x) = q(x'|x)A(x',x),
$$
where $q(x|x')$ is the probability density of the proposal distribution and $A(x',x)$ is the acceptance criteria (note that it is both a function of the current state and the updated state).
The accept-reject step consists of proposing a new sample,
$$
x' \sim Q(\cdot|x)
$$
where $Q(\cdot|x)$ is the proposal distribution, followed by checking if we accept the new sample.
We can now ask what acceptance criteria will ensure that the Markov Chain satisfies detailed balance with a stationary distribution $\pi(x)$.

::: {.callout title="Metropolis-Hastings Acceptance Criteria"}
$$
A(x',x) = \min \left(1, \frac{\pi(x')q(x| x')}{\pi(x)q(x' | x)}\right)
$$
where $q(x'|x)$ is probability density of the proposal distribution of $x'$ given a known $x$.
:::

The Metropolis-Hastings Algorithm is a way to construct a Markov Chain with known stationary distribution.
This algorithm breaks up the sampling procedure into two steps, the proposal step and the acceptance step.
By using an acceptance step that satisfies detailed balance, we can ensure that the stationary distribution of the algorithm is the probability distribution we want to sample from.
This is beneficial when you do not have another way to sample from the required distribution.

The full algorithm for Markov Chain Monte Carlo is:

1) Start with an initial sample $x$.
2) Generate a proposed sample $x' \sim Q(\cdot | x)$
3) Accept sample with probability $A(x',x)$.
4) Repeat steps 2-4 as desired.

This will generate a sequence of samples that have the stationary distribution $\pi(x)$.

```{r echo=T}
number_samples = 1000
initial_samples= c(1.0,5.0,-10.0,40.0)
mu = 40
sd = 2

transition_kernel <- function(current){
    proposed = current + rnorm(1, 0, 1)
    acceptance_criteria = dnorm(proposed, mu, sd)*dnorm(current, proposed, 1)/(dnorm(current, mu, sd)*dnorm(proposed, current, 1));
    if(runif(1,0,1) > acceptance_criteria){
        proposed = current
    }
    return(proposed)
}

generate_mcmc_chain <- function(initial_sample){
    return(purrr::accumulate(.x = 1:number_samples,
                      .f = ~ transition_kernel(.x),
                      .init = initial_sample))
}

purrr::map(.x = initial_samples,
           .f = ~ tibble::tibble(Sample = generate_mcmc_chain(.x)) |>
           dplyr::mutate(Iteration = dplyr::row_number())) |>
    dplyr::bind_rows(.id = "Chain") |>
    ggplot2::ggplot(ggplot2::aes(x = Iteration,y = Sample)) +
    ggplot2::geom_point() +
    ggplot2::facet_wrap(~Chain)
```

There are hyperparmeters within the above algorithm that should be understood, the most important of which are associated with the proposal distribution.
By calibrating the proposal distribution it is possible to end up with fast (or slower) convergence to the stationary distribution.

```{r echo=F}
number_samples = 200
initial_sample = 1
mu = 40
sd = 2
proposal_sds = 2^(seq(0,3))

transition_kernel <- function(current, proposal_sd){
    proposed = current + rnorm(1, 0, proposal_sd)
    acceptance_criteria = dnorm(proposed, mu, sd)*dnorm(current, proposed, proposal_sd)/(dnorm(current, mu, sd)*dnorm(proposed, current, proposal_sd));
    if(runif(1,0,1) > acceptance_criteria){
        proposed = current
    }
    return(proposed)
}

generate_mcmc_chain <- function(proposal_sd){
    return(purrr::accumulate(.x = 1:number_samples,
                      .f = ~ transition_kernel(.x, proposal_sd),
                      .init = initial_sample))
}

purrr::map(.x = proposal_sds,
           .f = ~ tibble::tibble(Sample = generate_mcmc_chain(.x)) |>
           dplyr::mutate(Iteration = dplyr::row_number(), sd = .x)) |>
    dplyr::bind_rows() |>
    ggplot2::ggplot(ggplot2::aes(x = Iteration,y = Sample)) +
    ggplot2::geom_point() +
    ggplot2::facet_wrap(~sd)
```

In our simple example the only hyperparameter within the proposal distribution is the proposal variance.
Figure XX demonstrates the effect of changing the proposal variance.
The time to reach the stationary distribution is obviously affected by the chosen variance and there will be an optimal choice (that is problem dependent).
It is important to note that as the proposal variance increases, we also observe an increased in the auto-correlation between samples.
This is partly attributed to an increased probability of the proposed sample being rejected.
Having a wider proposal distribution has made it more likely that samples are rejected.

::: {.callout-warning}
The implementation presented above is prone to numerical issues.
Numerical methods in statistics often have to deal with underflow and overflow, as such it is better to do computations in log-space.
:::

::: {.callout title="Proportionality"}
A very helpful property of the acceptance criteria is that we only need to know the stationary distribution up to a constant of proportionality.
Therefore, if we have a function,
$$
f(x) \propto \pi(x),
$$
we can still construct the Markov Chain with stationary distribution $\pi(x)$ by
$$
A(x',x) = \min \left(1, \frac{f(x')q(x| x')}{f(x)q(x' | x)}\right).
$$
:::

The majority of methods that speed up the convergence of Markov Chain Monte Carlo focus on choosing better proposal distributions.
A small and incomplete list of these methods include:

- Langevin Monte Carlo
- Hamiltonian Monte Carlo

## Autocorrelation of samples

### Limitations

So many many things.

